---
title: "Razorback Sucker Elemental Isotope Analysis"
author: "Alejandro Aragon"
date: "12/21/2021"
output:
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    code_folding: show
    number_sections: yes
    toc: yes
---

```{r, echo=FALSE}

knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, width = 100)
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, fig.width = 6)

```

# San Juan River Razorback Suckers

__Razorback Suckers__ were collected in 2014 on the San Juan River.
Elemental [isotopic ratios](https://en.wikipedia.org/wiki/Isotope_analysis)
from razorback suckers were analyzed for Ba (Barium 56), Ca (Calcium 20), Mg (Magnesium 12), and Sr (Strontium 38).
razorback suckers are non-lethally obtained and are used to detect natal origin since the
material in the suckers are partially developed early in life.

__The issue__ is that hatchery fish can get into the river and lose their tags.
It is important for environmental resource managers to know whether untagged fish are wild or hatchery fish.
There are five fish sources in the dataset.

```
5 Sources

Hatchery
  DEX = Dexter National Fish Hatchery
  GJH = Ouray National Fish Hatchery, Grand Valley Unit

Wild
  NAP = NAPI ponds
  SJR = San Juan River

Unknown
  UNK = untagged Razorback Suckers captured in the San Juan River
        these could be from any of the above sources
```

__Goal__ is to test whether the known source populations have different multivariate means, and if so, if it is possible to build a model that predicts the origin of razorback suckers.
  
# Data

## Summary

```{r}
sjrs.full <- read.csv("F:/Data Analysis/SanJuanRazorbackSuckers_data2014.csv")
str(sjrs.full)

summary(sjrs.full)
```

Adding `Source.Type` as a variable for a more general view of the sources. 

```
Source Type

HCH = Hatchery
    DEX = Dexter National Fish Hatchery
    GJH = Ouray  National Fish Hatchery, Grand Valley Unit

WLD = Wild 
    NAP = NAPI ponds
    SJR = San Juan River

Unknown
  UNK = untagged Razorback Suckers captured in the San Juan River
        these could be from any of the above sources
```

```{r}
sjrs.full <- na.omit(sjrs.full)
sjrs.full$Source.Type <- rep(NA, nrow(sjrs.full))
sjrs.full$Source.Type[ sjrs.full$Source %in% c("DEX", "GJH") ] <- "HCH"
sjrs.full$Source.Type[ sjrs.full$Source %in% c("NAP", "SJR") ] <- "WLD"
sjrs.full$Source.Type <- as.factor(sjrs.full$Source.Type)
```

## Preliminary Visualization

`Station` has 67 levels making a pairing table challenging to read. 

```{r, fig.height = 8, fig.width = 10}
library(ggplot2)

p <- ggplot(sjrs.full, aes(Station, fill = Source, color = Source.Type))
p <- p + geom_bar(stat = "count")
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + labs( title = "Station Count (Colored by Source)")
print(p)

```


Subsetting dataset to remove `Sort.Key`, `Station`,  and `Type`. Suckers from an unknown source is also removed here as it provides no information to the goal (identifying source by isotopes).

```{r}
sjrs <- subset(sjrs.full, subset = (Source != "UNK"), select = c(Source, Ba137:Source.Type))

str(sjrs)
summary(sjrs)
```

Subsetting data has removed any missing information.

```{r, fig.height = 8, fig.width = 8,  message = FALSE, warning = FALSE}
# Scatterplot matrix
library(ggplot2)
library(GGally)

p <- ggpairs(sjrs
            , mapping = ggplot2::aes(colour = Source.Type, alpha = 0.5)
            , upper = list(continuous = "density", combo = "box")
            , lower = list(continuous = "points", combo = "dot")
            #, lower = list(continuous = "cor")
            )
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + labs(title = "Known Fish Paired Plots (Colored by Source Type)"
              , subtitle = "Upper: Density / Lower: Scatter")

p1 <- ggpairs(sjrs
            , mapping = ggplot2::aes(colour = Source, alpha = 0.5)
            , upper = list(continuous = "density", combo = "box")
            , lower = list(continuous = "points", combo = "dot")
            #, lower = list(continuous = "cor")
            )
p1 <- p1 + theme(axis.text.x = element_text(angle = 90))
p1 <- p1 + labs(title = "Known Fish Paired Plots(Colored by Source)"
                , subtitle = "Upper: Density / Lower: Scatter")

print(p)
print(p1)
```



### Preliminary Scatter Plot Observations

* High correlation between among isotopes of the same element.
  + Dropping isotopes so one isotope of each element is represented in the model
  + a principle component of each element to be considered
* Barium isotopes would benefit from a transformation to normalize the distribution
* Calcium isotope has two observations that are skewing the distribution. 
* Strontium isotopes are indicative of razorback suckers from Dexter National Fish Hatchery
* Source Type adds no obvious information for preliminary inspection
  + Drop Source Type Variable

## Clean and Transforming data

__Transformations__

```{r}
sjrs$Ba137lg <- log10(sjrs$Ba137)
sjrs$Ba138lg <- log10(sjrs$Ba138)
```

__Restriction__

```{r}
sjrs <- sjrs[!(sjrs$Ca43 < .55),]
```

### Partitioning Data For Training

```{r}
library(caret)

sjrs$Source <- droplevels(sjrs$Source)

set.seed(1122)
trainIndex <- createDataPartition(sjrs$Source, p = .7, list = FALSE, times = 1)
train.dat <- sjrs[ trainIndex,]
test.dat <- sjrs[-trainIndex,]
```

__Train Data Visualization (After Transformations)__

Unnecessary step, only done to confirm log transformation is appropriate for training data.

```{r, fig.height = 8, fig.width = 8,  message = FALSE, warning = FALSE, echo=FALSE}
library(ggplot2)
library(GGally)
p <- ggpairs(subset(train.dat, select = c(Source, Ba138lg, Ca43, Mg24, Sr87))
            , mapping = ggplot2::aes(colour = Source, alpha = 0.5)
            , upper = list(continuous = "density", combo = "box")
            , lower = list(continuous = "points", combo = "dot")
            #, lower = list(continuous = "cor")
            )
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + labs(title = "Known Fish Paired Plots (Colored by Source)"
              , subtitle = "Upper: Density / Lower: Scatter")
print(p)
```

## Data Summary

Razorback suckers from the Dexter National Fish Hatchery can be easily identified by most combinations of the predictor variables. Ba138 in combination with any other predictor variable allows suckers from the Ouray National Fish Hatchery, Grand Valley Unit to be visually categorized. 

# Multivariate Analysis of Variance (MANOVA)

### Assumptions

Shapiro-Wilk test for multivariate normality, as well as QQ-plots comparing the Mahalanobis D2 distance to a chi-squared distribution.

__Test Hypothesis__

Null hypothesis: The data is normally distributed. 

Alternative hypothesis: Data is not normally distributed.

If $p$-value < .05 reject $H_0$.


```{r}
library(mvnormtest)

iso.vars <- c("Ca43", "Mg24", "Mg25", "Mg26", "Sr86"
              , "Sr87", "Sr88", "Ba137lg", "Ba138lg" )

# The data needs to be transposed t() so each variable is a row with observations as columns.
sw.dex <- mshapiro.test(t(subset(train.dat, subset = (Source == "DEX"), select = iso.vars)))
sw.gjh <- mshapiro.test(t(subset(train.dat, subset = (Source == "GJH"), select = iso.vars)))
sw.nap <- mshapiro.test(t(subset(train.dat, subset = (Source == "NAP"), select = iso.vars)))
sw.sjr <- mshapiro.test(t(subset(train.dat, subset = (Source == "SJR"), select = iso.vars)))
sw.results <- list(sw.dex, sw.gjh, sw.nap, sw.sjr)
```

```{r, fig.height = 8, fig.width = 8}
library(mvnormtest)

# Graphical Assessment of Multivariate Normality
f.mnv.norm.qqplot <- function(x, name = "") {
  x <- as.matrix(x)     
  center <- colMeans(x) 
  n <- nrow(x);
  p <- ncol(x);
  cov <- cov(x);
  d <- mahalanobis(x, center, cov) # distances
  qqplot(qchisq(ppoints(n), df=p), d
    , main=paste("QQ Plot MV Normality:", name)
    , ylab="Mahalanobis D2 distance"
    , xlab="Chi-squared quantiles")
  abline(a = 0, b = 1, col = "red")
}

par(mfrow=c(2,2))
f.mnv.norm.qqplot(subset(train.dat, subset = (Source == "DEX"), select = iso.vars), "DEX")
f.mnv.norm.qqplot(subset(train.dat, subset = (Source == "GJH"), select = iso.vars), "GJH")
f.mnv.norm.qqplot(subset(train.dat, subset = (Source == "NAP"), select = iso.vars), "NAP")
f.mnv.norm.qqplot(subset(train.dat, subset = (Source == "SJR"), select = iso.vars), "SJR")
```

__Results__

$p$-value for the Shapiro-Wilk test for Dexter National Fish Hatchery (DEX) is `r signif(sw.dex$p.value,6)`. $`r signif(as.numeric(unlist(sw.dex)["p.value"]),6)` < .05$, reject $H_0$ and conclude that the observations for DEX are not normally distributed.

$p$-value for the Shapiro-Wilk test for Ouray  National Fish Hatchery, Grand Valley Unit (GJH) is `r signif(as.numeric(unlist(sw.gjh)["p.value"]),6)`. $`r signif(as.numeric(unlist(sw.gjh)["p.value"]),6)` < .05$, fail to reject $H_0$ and conclude that the observations for GJH are not normally distributed.

$p$-value for the Shapiro-Wilk test for NAPI ponds (NAP) is `r signif(as.numeric(unlist(sw.nap)["p.value"]),6)`. $`r signif(as.numeric(unlist(sw.nap)["p.value"]),6)` < .05$ fail to reject $H_0$ and conclude that the observations for NAP are not normally distributed.

$p$-value for the Shapiro-Wilk test for San Juan River (SJR) is `r signif(as.numeric(unlist(sw.sjr)["p.value"]),6)`. $`r signif(as.numeric(unlist(sw.sjr)["p.value"]),6)` < .05$ reject $H_0$ and conclude that the observations for SJR are not normally distributed.

__Conclusion on Assumption of Normality__

Since the assumption of normality is not met a non-parametric measure of MANOVA will need to be done.

## MANOVA Model

MANOVA will be measured by the Hotelling-Lawley trace test.

### Comparison Across Sources

__Test Hypothesis__

$$
H_0: \mu_{DEX} = \mu_{GJH} = \mu_{NAP} = \mu_{SJR}  
\textrm{  versus  }
H_\alpha: \mu_{DEX} \ne \mu_{GJH} \ne \mu_{NAP} \ne \mu_{SJR}
$$

Null hypothesis: The fish from the four sources are the same across isotopes.

Alternative hypothesis: The fish from the four sources differ in isotopes.

```{r}
library(car)

lm.man <- lm(cbind(Ca43, Mg24, Mg25, Mg26, Sr86, Sr87, Sr88, Ba137lg, Ba138lg) 
             ~ Source, data = train.dat)
man.train <- Manova(lm.man, test = "Hotelling-Lawley")
man.train
```

__Results__

The results from the Hotelling-Lawley test show that we have sufficient evidence to reject the null hypothesis and conclude that at least one of the sources differs from the other.


### Multiple Comparisons (Between Sources)

__Test Hypothesis__


$$
H_0: \mu_{i} = \mu_{j}  
\textrm{  versus  }
H_\alpha: \mu_{i} \ne \mu_{j} 
$$

Null hypothesis: mean of `source i` is equal to the mean of `source j`. 

Alternative hypothesis: mean of `source i` is not equal to the mean of `source j`.

This test is repeated for all sources so that the mean of each source has been tested against each of the others.

```{r}
library(car)
# Multivariate MANOVA test

Source.list <- sort(unique(train.dat$Source))

for (i in 1:(length(Source.list) - 1)) {
  for (j in (i + 1):length(Source.list)) {
    # print a header to indicate which comparisons are being made
    cat("\n\n")
    cat(paste("***** Comparison between", i, Source.list[i], "and", j, Source.list[j]))

    # perform pairwise comparison
    
    man.pair <- Manova(lm(cbind(Ba137, Ba138, Ca43, Mg24, Mg25, Mg26, Sr86, Sr87, Sr88)
        ~ Source, data = subset(train.dat, (Source %in% Source.list[c(i, j)]))
           ))
    # print result
    print(man.pair)
  }
}
```

__Results__

At $\alpha = .05$ reject $H_0$ and conclude the means between the following pairs are not equal:
  DEX and GJH
  DEX and NAP
  DEX and SJR
  GJH and NAP
  and
  GJH and SJR

Fail to reject the null and conclude that the means of NAP and SJR are equal.

## Canonical Discriminant Functions (Further Visualizations)

The canonical discriminant analysis will indicate the directions that provide the greatest ability to distinguish between the groups.
  
```{r}
library(candisc)

can.train <- candisc(lm.man)
summary(can.train)
```

```{r, fig.height = 8, fig.width = 8,  message = FALSE, warning = FALSE, echo=FALSE}
library(ggplot2)
library(GGally)

p <- ggpairs(can.train$scores
            , mapping = ggplot2::aes(colour = Source, alpha = 0.5)
            , upper = list(continuous = "density", combo = "box")
            #, lower = list(continuous = "points", combo = "dot")
            , lower = list(continuous = "cor")
            , title = "Canonical discriminant variables by source"
            )
print(p)
```


__Results__

The discriminant variables all except $Can3$ show some sort of difference between the sources because of their distributions but $Can1$ shows the clearest differences and similarities making it the most useful.

## MANOVA Final Results

Most of the sources can be differentiated by the mean of the isotopes. Razorback suckers that lose their tags can be traced to their source by analyzing the elemental isotopes in the fish.

# Cluster Analysis


__Goal__ is to find a clustering method that creates a dendrogram that seems to have a relatively small number of clusters that are different between clusters but similar within clusters.

### Linkage Method Selection

This section is unnecessary in understanding the research question. The cost to run this code is more expensive than its benefit and it's included here for future analysis and ease of examining the efficacy of other linkage methods.

```{r, eval=FALSE}
# eval=FALSE so this chunk doesn't evaluate 

wd.clust <- NbClust(train.num, method = "ward.D", index = "all")
wd.clust$Best.nc

wd2.clust <- NbClust(train.num, method = "ward.D2", index = "all")
wd2.clust$Best.nc

sgl.clust <- NbClust(train.num, method = "single", index = "all")
sgl.clust$Best.nc

com.clust <- NbClust(train.num, method = "complete", index = "all")
com.clust$Best.nc

aveclust <- NbClust(train.num, method = "average", index = "all")
aveclust$Best.nc

mcq.clust <- NbClust(train.num, method = "mcquitty", index = "all")
mcq.clust$Best.nc

med.clust <- NbClust(train.num, method = "median", index = "all")
med.clust$Best.nc

cen.clust <- NbClust(train.num, method = "centroid", index = "all")
cen.clust$Best.nc
```

## Average Linkage Method


```{r message=FALSE, warning=FALSE}
library(NbClust)

# Removing factor variables
train.sub <- subset(train.dat, select = iso.vars)
train.num <- as.numeric(as.matrix(train.sub))

aveclust <- NbClust(train.num, method = "average", index = "all")

aveclust$Best.nc
```


## Plotting

### Dendrogram

```{r}
# create distance matrix between points
train.dist <- dist(train.sub)
# create dendrogram
hc.complete  <- hclust(train.dist, method = "average")

# create a column with group membership
train.dat$cut.comp <- factor(cutree(hc.complete, k = 3))
```


```{r, fig.height = 5, fig.width = 8}
plot(hc.complete
     , hang = -1
     , main = "Average Linkage Method Using 3 Clusters"
#     , labels = train.sub[,1]
     )
rect.hclust(hc.complete, k = 3)
```

The density of the dendrogram makes this plot hard to understand. Plotting against principal components gives a better visual understanding of the classifications.

### PCA

```{r, fig.height = 5, fig.width = 8}
library(cluster)

clusplot(train.dat, cutree(hc.complete, k = 3)
        , color = TRUE, labels = 0, lines = 0
        , cex = 2, cex.txt = 1, col.txt = "gray20"
        , main = "Average Method using 3 clusters"
        , sub = NULL
        , col.p = train.dat$cut.comp)
```

## Cluster Analysis Results

### Visualizing Efficacy

__Efficacy By Source__

```{r, fig.height = 8, fig.width = 10, echo=FALSE}
library(ggplot2)

# plot data by Source with clusters indicated

p1 <- ggplot(train.dat, aes(x = Ba137lg, y = Sr86, colour = cut.comp))
p1 <- p1 + geom_point(size = 2)
p1 <- p1 + labs(title = "By source with clusters indicated")
p1 <- p1 + facet_wrap( ~ Source, nrow = 1)


# plot data by cluster with Source indicated
p2 <- ggplot(train.dat, aes(x = Ba137lg, y = Sr86, colour = Source))
p2 <- p2 + geom_point(size = 2)
p2 <- p2 + labs(title = "By cluster with source indicated")
p2 <- p2 + facet_wrap( ~ cut.comp, nrow = 1)

library(gridExtra)
grid.arrange(grobs = list(p1, p2), ncol=1
             , top = paste("Clustering Using Average Linkage With Three Clusters"))
```

__Efficacy By Source Type__

```{r, fig.height = 8, fig.width = 10, echo=FALSE}
library(ggplot2)

# plot data by Source with clusters indicated

p1 <- ggplot(train.dat, aes(x = Ba137lg, y = Sr86, colour = cut.comp))
p1 <- p1 + geom_point(size = 2)
p1 <- p1 + labs(title = "By source type with clusters indicated")
p1 <- p1 + facet_wrap( ~ Source.Type, nrow = 1)


# plot data by cluster with Source indicated
p2 <- ggplot(train.dat, aes(x = Ba137lg, y = Sr86, colour = Source.Type))
p2 <- p2 + geom_point(size = 2)
p2 <- p2 + labs(title = "By cluster with source type indicated")
p2 <- p2 + facet_wrap( ~ cut.comp, nrow = 1)

library(gridExtra)
grid.arrange(grobs = list(p1, p2), ncol=1
             , top = paste("Clustering Using Average Linkage With Three Clusters"))
```


The plots show there would be clear errors in classification of the origin of the razorback suckers particularly when trying to differentiate fish sourced from NAPI ponds and the San Juan River. This analysis is not unexpected given the results of the MANOVA analysis. 

# Prediction

## Principal Components Analysis (PCA)

The model can be simplified without sacrificing information gain by taking a principal component of correlated isotopes.

The correlation can be seen on previous pairs plots but will be shown again here.

```{r, fig.height = 5, fig.width = 5}
library(ggplot2)
library(GGally)

train.x <- subset(train.dat, select = iso.vars)

p <- ggcorr(train.x, method = c("everything", "pearson"))
p <- p + labs(title = "Correlation Plot")

print(p)
```

__Barium Isotope Variables__:

```{r}
pca.Ba <- princomp( ~ Ba137lg + Ba138lg, data = train.dat, cor = FALSE)
train.dat$PC1.Ba <- pca.Ba$scores[,"Comp.1"]

summary(pca.Ba)
print(loadings(pca.Ba), cutoff = 0)
```

```{r}
# Adding the PC to the test dataset
pca.Ba <- princomp( ~ Ba137lg + Ba138lg, data = test.dat, cor = FALSE)
test.dat$PC1.Ba <- pca.Ba$scores[,"Comp.1"]
```


`Comp.1` explains `r signif(100*summary(pca.Ba)$sdev[1]^2/sum(summary(pca.Ba)$sdev^2), 5)`% of the variability of both the Barium isotopes.

__Calcium Isotope Variables__:

```{r}
pca.Ca <- princomp( ~ Ca43, data = train.dat, cor = FALSE)
train.dat$PC1.Ca <- pca.Ca$scores[,"Comp.1"]

summary(pca.Ca)
print(loadings(pca.Ca), cutoff = 0)
```

`Comp.1` explains `r signif(100*summary(pca.Ca)$sdev[1]^2/sum(summary(pca.Ca)$sdev^2), 5)`% of the variability of the Calcium isotope.

__Magnesium Isotope Variables__:

```{r}
pca.Mg <- princomp( ~ Mg24 + Mg25 + Mg26, data = train.dat, cor = FALSE)
summary(pca.Mg)
print(loadings(pca.Mg), cutoff = 0)

train.dat$PC1.Mg <- pca.Mg$scores[,"Comp.1"]
```

Note that `Comp.1` explains `r signif(100*summary(pca.Mg)$sdev[1]^2/sum(summary(pca.Mg)$sdev^2), 5)`% of the variability of all three of the Magnesium isotopes.

__Strontium Isotope Variables__:

```{r}
pca.Sr <- princomp( ~ Sr86 + Sr87 + Sr88, data = train.dat, cor = FALSE)
summary(pca.Sr)
print(loadings(pca.Sr), cutoff = 0)

train.dat$PC1.Sr <- -pca.Sr$scores[,"Comp.1"]
```

Note that `Comp.1` explains `r signif(100*summary(pca.Sr)$sdev[1]^2/sum(summary(pca.Sr)$sdev^2), 5)`% of the variability of all three of the Strontium isotopes.

```{r, fig.height = 8, fig.width = 8,  message = FALSE, warning = FALSE, echo=FALSE}
library(ggplot2)
library(GGally)

p <- ggpairs(
             subset(train.dat, select = c(PC1.Sr, PC1.Mg, PC1.Ba, PC1.Ca, Source.Type) )
            , mapping = ggplot2::aes(colour = Source.Type, alpha = 0.5)
            , upper = list(continuous = "density", combo = "box")
            , lower = list(continuous = "points", combo = "dot")
            )
print(p)


```

```{r}

# Adding the PC to the test dataset
pca.Ba <- princomp( ~ Ba137lg + Ba138lg, data = test.dat, cor = FALSE)
test.dat$PC1.Ba <- pca.Ba$scores[,"Comp.1"]

pca.Ca <- princomp( ~ Ca43, data = test.dat, cor = FALSE)
test.dat$PC1.Ca <- pca.Ca$scores[,"Comp.1"]

pca.Mg <- princomp( ~ Mg24 + Mg25 + Mg26, data = test.dat, cor = FALSE)
test.dat$PC1.Mg <- pca.Mg$scores[,"Comp.1"]

pca.Sr <- princomp( ~ Sr86 + Sr87 + Sr88, data = test.dat, cor = FALSE)
test.dat$PC1.Sr <- -pca.Sr$scores[,"Comp.1"]
```

## Regression Modeling

From visual inspection predicting hatchery and wild suckers will not be difficult so, both a binomial regression (fitted against `Source.Type`), and a multinomial regression (fitted on `Source`) will be done.

### Binomial Model

```{r}
library(nnet)

Bi.fit <- multinom(Source.Type ~ PC1.Ba + PC1.Ca + PC1.Mg + PC1.Sr, data = train.dat)
summary(Bi.fit)
```


```{r}
# Predicting the values for train dataset
test.dat$BiPred <- predict(Bi.fit, newdata = test.dat, "class")

# Building classification table
tab <- table(test.dat$Source.Type, test.dat$BiPred)
tab

# Calculating accuracy - sum of diagonal elements divided by total obs
round((sum(diag(tab))/sum(tab))*100,2)
```

#### Binomial Model Reduction

```{r}
Bi.fit.red.AIC <- step(Bi.fit, direction="both", trace = 0)

Bi.fit.red.AIC$anova
summary(Bi.fit.red.AIC)
```

```{r}
# Predicting the values for train dataset
test.dat$BiPred <- predict(Bi.fit.red.AIC, newdata = test.dat, "class")

# Building classification table
tab <- table(test.dat$Source.Type, test.dat$BiPred)
tab

# Calculating accuracy - sum of diagonal elements divided by total obs
round((sum(diag(tab))/sum(tab))*100,2)
```

As expected the model accuracy in predicting `Source.Type` is 100%.

### Multinomial Model

```{r}
library(nnet)

mult.fit <- multinom(Source ~ PC1.Ba + PC1.Ca + PC1.Mg + PC1.Sr, data = train.dat)
summary(mult.fit)
```

```{r}
# Predicting the values for train dataset
test.dat$MultPred <- predict(mult.fit, newdata = test.dat, "class")

# Building classification table
tab <- table(test.dat$Source, test.dat$MultPred)
tab

# Calculating accuracy - sum of diagonal elements divided by total obs
round((sum(diag(tab))/sum(tab))*100,2)
```

#### Multinomial Model Reduction

```{r}
mult.fit.red.AIC <- step(mult.fit, direction="both", trace = 0)

mult.fit.red.AIC$anova
summary(mult.fit.red.AIC)
```

```{r}
# Predicting the values for train dataset
test.dat$MultPred <- predict(mult.fit.red.AIC, newdata = test.dat, "class")

# Building classification table
tab <- table(test.dat$Source, test.dat$MultPred)
#tab

# Calculating accuracy - sum of diagonal elements divided by total obs
round((sum(diag(tab))/sum(tab))*100,2)
```

The final multinomial model contains `PC1.Ba` and `PC1.Sr` as predictor variables and has an accuracy rate of `r round((sum(diag(tab))/sum(tab))*100,2)`%. The error is in predicting suckers originating from NAPI ponds and the San Juan River which is expected based on previous analysis.

## Model Efficacy

```{r}
library(caret)

conf.mat <- confusionMatrix(test.dat$MultPred, test.dat$Source)
conf.mat
```


# Conclusion

Wild razorback suckers and those from hatcheries can be differentiated by their elemental isotopic ratios and their origin can be predicted to 100% accuracy. Determining the specific source of the suckers is more difficult, with a final model accuracy of 84.39%. Fish from the Dexter National Fish Hatchery and Ouray National Fish Hatchery, Grand Valley Unit are easily differentiated from each other and those from the NAPI ponds and San Juan River, however, differentiating between fish from NAPI ponds and the San Juan River is not possible. The elemental isotopic ratio that has the most weight in classification of origin are the Barium isotopes, which was shown to be significant in classifying both hatchery vs wild fish and fish by their source.

Similarities in elemental isotopic ratios can be explained by proximity:

```
Ouray National Fish Hatchery: Northeaster Utah
Dexter National Fish Hatchery: Southeastern New Mexico
NAPI ponds: Northwestern New Mexico
San Juan River: spans from Southern Utah to Northern New Mexico
```
NAPI ponds and the San Juan River at a point are within 4 miles of each other while they are approximately 300 miles away from the Ouray and Dexter National Fish Hatcheries, which are 900 miles away from each other. Similarities elemental isotopic ratios can be explained by proximity.

Overall, tracking hatchery sourced razorback suckers is possible for biologists and conservationists to do and I assume it would be possible for other fish that are sourced from hatcheries that are not in close proximity of the natural populations they supplement.


# Additional Reading

* Navajo Agricultural Products Industry (NAPI) [2014 report](https://www.fws.gov/southwest/sjrip/pdf/DOC_NAPI_Razorback_Ponds_2014_Final.pdf)
* Center for Biological Diversity [species notes](https://www.biologicaldiversity.org/species/fish/razorback_sucker/index.html)
* The Southwestern Native Aquatic Resources & Recovery Center (formerly the Dexter National Fish Hatchery & Technology Center) [reference page](https://www.fws.gov/southwest/fisheries/dexter/)
* The Colorado River Fishery Project [page](https://www.fws.gov/mountain-prairie/fisheries/coloradoRiver.php)






















